Statically typed Web programming

### 2012-12-10 papers/ecoop2013/repairing.tex
Ready for j-m

JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If
developers want to use another numeric data type, \eg{} representing integer values or arbitrary length values, they
can’t use the native arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because
the same concepts (\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the
code is also harder to write when it comes to mix different types of numeric values. Indeed, this requires to
manually take care of their compatibility: before adding an arbitrary length value with a \code{Number} value,
developers have to convert the latter in the former’s type.

@@

The following listing demonstrates the homogeneous syntax to perform an addition on different data types and the
operand coercion between heterogeneous types:

\begin{lstlisting}
def main(i: Rep[Int], j: Rep[Int],
         x: Rep[Double], y: Rep[Double]) = {
  println(i + j)
  // i is promoted to Rep[Double], m has type Rep[Double]
  val m = i + y
  // j is promoted to Rep[Double], n has type Rep[Double]
  val n = x + j
  println(m + n)
}
\end{lstlisting}

@@

(TODO Example)

On the other side of the coin, because that is not possible to define a library without defining how to translate it
on both server and client sides, there is no simple way to let DSLs users define their own libraries. This feature
comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the
server-side program from the initial program source (that contains data type definitions such as classes
definitions). In our case, the client-side program and the server-side program are derived from the intermediate
representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
a value of type \code{Rep[A]}). Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is 
limited to record types and provides only projection functions, so for more complex cases users have to write some
repetitive and boilerplate code.

@@

However, the next release of Scala will have support for macros making it possible to generate reified data types
from a concise syntax without relying on a compiler extension.

@@

Our DSL aims to give developers a safer and more expressive way to write the client-side code. We focused on the main
tasks of RIAs development: handling events, asynchronous programming and DOM fragments definition, as well as on
more general programming concerns: \code{null} references handling and type safety.

@@

Using our DSL, we were able to divide by two the length of some parts of code (DOM definition and \code{null}
reference handling). The size of code parts dealing with events did not change but they gain type safety. Finally,
code handling asynchronous computations were easier to modularize.

@@

\begin{table}
\centering
\begin{tabular}{| l | c | c | c | c | c | c | c | c |}
\hline
& jQuery & TypeScript & Roy & GWT & Kotlin & Opa & Dart & js-scala \\
\hline
Event handling & * & & & *** & & ? & & *** \\
\hline
DOM definition & * & & & *** & ** & *** & * & *** \\
\hline
Asynchronous programming & ** & & *** & & & *** & * & *** \\
\hline
Null references & ** & & *** & & * & ? & & *** \\
\hline
Type coercion & & & & * & * & ? & * & *** \\
\hline
General programming language features & * & ** & ** & * & *** & ? & * & *** \\
\hline
Static typing & & * & ** & ** & ** & ** & * & ** \\
\hline
Easiness for defining libraries & * & *** & *** & *** & *** & *** & *** & * \\
\hline
Client-server code sharing & & * & * & * & * & ** & * & *** \\
\hline
\end{tabular}
\caption{Support level of Web programming features by famous tools. Zero star means that the feature is not supported
at all, three stars mean a strong support.}
\label{comparison}
\end{table}

@@

GWT exposes the browser’s APIs through type safe Java APIs and simplifies DOM fragments definition. However it still
uses a callback-driven programming models for asynchronous computations.

@@

Opa is a language specifically designed for Web programming.

Dart is another language designed specifically for Web programming.

@@

Writing RIAs is challenging because the JavaScript language and the Web browser’s native APIs have several
inadequacies for this purpose.

We provide a way to write statically typed client-side code, we simplify asynchronous programming

We provide a safer and more expressive language than JavaScript and its native APIs to write RIAs.

More code sharing.

>>>>

JavaScript comes with only one numeric type, \code{Number}, that represents double-precision 64-bit values. If
developers want to use another numeric data type, \eg{} representing rational numbers, they can’t use the native
arithmetic operators (\code{+}, \code{-}, \etc) on them, making the code less readable because the same concepts
(\eg{} adding two numeric values) are described by different operators or methods. Furthermore, the code is also
harder to write when it comes to mix different types of numeric values. Indeed, this requires to manually take care
of their compatibility: before adding a rational number with a \code{Number} value, developers have to convert the
former in the latter’s type. Consider as an example the listing \ref{type-coercion} mixing rational numbers and
native numbers.

\begin{lstlisting}[language=JavaScript,label=type-coercion,caption=Mixing integer and floating-point arithmetic in
JavaScript.]
// i and j are rational numbers, x and y are native numbers
var main = function (i, j, x, y) {
  console.log(rat_plus(i, j));
  var m = rat_toNumber(i) + y;
  var n = x + rat_toNumber(j);
  console.log(m + n);
};
\end{lstlisting}

@@

The listing \ref{type-coercion} can be rewritten as shown in listing \ref{type-coercion-solution} with our DSL.
The \code{+} operator can be safely used to add rational and floating-point numbers and type promotion is
automatically handled by the type system. It is worth noting that the \code{Rational} data type can be written
independently of the type coercion system.

\begin{lstlisting}[label=type-coercion-solution,caption=Type-coercion automatically handled by our DSL]
def main(i: Rep[Rational], j: Rep[Rational],
         x: Rep[Double], y: Rep[Double]) = {
  println(i + j)
  // i is promoted to Rep[Double], m has type Rep[Double]
  val m = i + y
  // j is promoted to Rep[Double], n has type Rep[Double]
  val n = x + j
  println(m + n)
}
\end{lstlisting}

@@

On the other side of the coin, because that is not possible to define a library without defining how to translate it
on both server and client sides, there is no simple way to let DSLs users define their own libraries. This feature
comes almost for free with approaches like GWT, Fay or Kotlin because they derive the client-side program and the
server-side program from the initial program source (that contains data type definitions such as classes
definitions). In our case, the client-side program and the server-side program are derived from the intermediate
representation produced by the evaluation of the initial program. So, only concepts reified in an intermediate
representation during this evaluation can be part of the final programs. However regular Scala’s classes definition
are not reified in an intermediate representation (it would mean that the constructor of a type \code{A} would yield
a value of type \code{Rep[A]}). Scala-virtualized~\cite{Moors12_Virtualized} provides such a mechanism but is 
limited to record types and provides only projection functions, so for more complex cases users have to write some
repetitive and boilerplate code. (FIXME Example?)

@@

However, the next release of Scala will have support for macros making it possible to generate reified data types
from a concise syntax without even relying on a compiler extension such as Scala-virtualized.

@@

Our goal was to define, with a minimal effort, a language for RIAs development. The main challange was to provide
language constructs both safer and more expressive without abstraction penalty. We focused on the main concerns of
RIA development: handling events, asynchronous programming and DOM fragments definition, as well as on more general
programming concerns: \code{null} references handling and type safety.

@@

Using our DSL, we were able to completely rewrite the client-side part of our application, and divided by two the
length of some parts of code (DOM definition and \code{null} reference handling). The size of code parts dealing with
events did not change but gained type safety. Finally, code handling asynchronous computations were easier to
modularize. An interesting side-effect of the design of our DSL as a compiled embedded DSL is the ability to reuse
some code between both client and server sides (DOM definition).

@@

\begin{table}
\centering
\begin{tabular}{| l | c | c | c | c | c | c | c | c |}
\hline
& jQuery & TypeScript & Roy & GWT & Kotlin & Opa & Dart & js-scala \\
\hline
Type safe event handling & & & & *** & & & & *** \\
\hline
DOM definition & * & & & *** & ** & *** & * & *** \\
\hline
Asynchronous programming & ** & & *** & & & *** & ** & *** \\
\hline
Null references & ** & & *** & & * & ** & & *** \\
\hline
Type coercion & & & & * & * & * & * & *** \\
\hline
%General programming language features & * & ** & ** & * & *** & ? & * & *** \\
%\hline
Static typing & & * & ** & ** & ** & ** & * & ** \\
\hline
Easiness for defining libraries & * & *** & *** & *** & *** & *** & *** & * \\
\hline
Client-server code sharing & & * & * & * & * & *  & * & *** \\
\hline
\end{tabular}
\caption{Support level of Web programming features by famous tools. Zero star means that the feature is not supported
at all, three stars mean a strong support.}
\label{comparison}
\end{table}

@@

GWT compiles Java code to JavaScript. It exposes the browser’s APIs through type safe Java APIs and simplifies DOM
fragments definition. However, it still uses a callback-driven programming models for asynchronous computations.

@@

Opa is a language and a full stack designed for Web programming. It features static typing with global
type inference, modules and pattern matching. The language provides a syntactic sugar for writing HTML fragments in
the code. The server and client parts of an application can be written in the Opa language, the system can decide
whether a code fragment will run on server-side, client-side or both.

Dart is a language designed specifically for Web programming. It has been designed to have a familiar syntax for
JavaScript developers and supports soft typing, classes and a module system. It also provides a high-level concurrent
programming system similar to an actor system. Finally, Dart code can be executed on both client and server sides.

Concerning the client-server code sharing feature, because our solution is the only one that gives full control on
the generated code on both client and server for a given piece of shared code, we considered that our support was
better than other approaches.

@@

Writing RIAs is challenging because the JavaScript language and the Web browser’s native APIs have several
inadequacies, making the code fragile, hard to reason about and to maintain.

We improved the js-scala DSL by adding first-class support for RIA development concerns such as events handling,
DOM fragments definition and asynchronous programming. Our work provides a safer and more expressive language than
JavaScript and its native APIs to write RIAs.

Our approach’s distinguishing support of code sharing between server and clients let us consider the exploration of
this feature more deeply in our future work, in order to factor out an increasing part of code between client and
servers in RIAs.
